---
title: "R Notebook - Bact-to-Batch PUBLICATION"
output:
  html_document:
    df_print: paged
---


A new method based on microbiota fingerprinting to guide animal allotments for animal experiments 


```{r load , echo=FALSE, warning=FALSE, message=FALSE}

print("R version :  version.string R version 4.1.0 (2021-05-18)")

print("Load packages")
library(phyloseq) #1.38.0
library(ggplot2) #3.4.0
library(stringr) #1.4.0"
library(dplyr) #1.0.9
library(GUniFrac)  #1.7
library(vegan) #2.6.2
library(svglite) #2.0.0
library(tidyverse) #1.3.1
library(grid) #4.1.0
library(ggpattern) #1.0.1
library(gridExtra) #2.3
library(kableExtra) #1.3.4
library(RcppAlgos) #2.6.0
library(analogue) #0.17-6
library(foreach) #1.5.2
library(anticlust) #O.6.2



normalize_custom <- function(dataset){
  normalized = (dataset -  min(dataset)) / (max(dataset) - min(dataset));
  return(normalized)
}


```

# NAIVE APPROACH - CONSTAT


## CONSTAT - All solutions by ranking score


```{r alldesign, echo=FALSE, warning=FALSE}

print("Load the data.frame containing the 92378 results - Allotment 20 samples 2 groups")

load("../Robjects/OUTPUT_TAB_STATS_COMBINATION_20_ALLOTMENT_WUNIFRAC_PUBLICATION_v.rdata1")

#Order by ranking score
tab_stats_combinaison_sorted =tab_stats_combinaison_ALL_WUNIFRAC[ order(tab_stats_combinaison_ALL_WUNIFRAC$METRIC_PVALUE),]

#Categorize the design according to quantile of ranking score 
quantile_metric= quantile(tab_stats_combinaison_sorted$METRIC_PVALUE, probs = c(0.05,0.95))
tab_stats_combinaison_sorted$group = "SUB-OPTIMAL DESIGN"
tab_stats_combinaison_sorted$group[tab_stats_combinaison_sorted$METRIC_PVALUE <= as.numeric(quantile_metric[1]) ] = "OPTIMAL DESIGN"
tab_stats_combinaison_sorted$group[tab_stats_combinaison_sorted$METRIC_PVALUE >= as.numeric(quantile_metric[2]) ] = "HIGH-RISK DESIGN"

#build the whole histogramm for the 92378 possible designs 
compare_mean <- tab_stats_combinaison_sorted %>%
        group_by(group) %>%
        summarise(Mean = mean(METRIC_PVALUE))
output_histo_allotement <- ggplot(tab_stats_combinaison_sorted, aes(x = METRIC_PVALUE, fill = group)) +
        geom_histogram(binwidth = 20, alpha = .5, position = "identity")  + scale_fill_manual(values=c("red", "green", "orange")) + labs(x = "RANKING SCORE") +  theme( panel.background = element_rect(fill = "white", colour = "black"), panel.grid.major=element_line(colour = "grey"))

output_histo_allotement

tab_stats_combinaison_sorted$group_char = tab_stats_combinaison_sorted$group

```

## CONSTAT -  MAX HIGH-RISK RANKING SCORE FROM ALL POSSIBLE DESIGNS

```{r, echo=FALSE, warning=FALSE}

tab_high_rank = tab_stats_combinaison_sorted[ tab_stats_combinaison_sorted$METRIC_PVALUE == max( tab_stats_combinaison_sorted$METRIC_PVALUE ), ]

  
###GROUP WORST
  
group_high_rank = str_split(tab_high_rank$GROUP,pattern = "")[[1]]

load("../Robjects/MOD.HIGH.RANK.ALLOTMENT.rdata")
plot(mod.high.rank, hull = FALSE, ellipse = TRUE, main=c(paste("Axis1Axis2, PERMANOVA p-value = ",tab_high_rank$betadiv_pvaleur,sep="")))


#fix css 
Theme_publi = theme(
 panel.background = element_rect(fill = "white", colour = "black"),
 axis.text.y = element_text(size = 14),
  axis.text.x = element_text(size = 14),
  panel.grid.major=element_line(colour = "grey"),
 legend.title=element_text(size=14),
 legend.text=element_text(size=12),
 axis.title.y=element_text(size=16, angle=90),
 axis.title.x=element_text(size=16),title = element_text(size=14))

load("../Robjects/PHY.RAREFIED.HIGH.RANK.ALLOTMENT.PLOT.rdata")

#alpha-diversity plot
al1<-ggplot(phy.rarefied.high.rank.plot, aes(x=Group, y=Richness, color=Group)) +
	  geom_boxplot() + geom_point(size = 2) + ggtitle(paste("Chao1, p-value = ",round(tab_high_rank$richness_pvaleur,3),sep="")) + Theme_publi
	al2<-ggplot(phy.rarefied.high.rank.plot, aes(x=Group, y=Evenness, color=Group)) +
	  geom_boxplot() + geom_point(size = 2) + ggtitle(paste("Shannon, p-value = ",round(tab_high_rank$eveness_pvaleur,3),sep="")) + Theme_publi
cpubli_high_rank <- cowplot::plot_grid(al1, al2, nrow = 1, ncol = 2, scale = .9, label_size = 28)
cpubli_high_rank


```

## CONSTAT - BEST RANKING SCORE FROM ALL POSSIBLE DESIGNS

Visualize the best ranking score. 

```{r, echo=FALSE, warning=FALSE}
tab_best = tab_stats_combinaison_sorted[ tab_stats_combinaison_sorted$METRIC_PVALUE == min( tab_stats_combinaison_sorted$METRIC_PVALUE ), ]
	
###GROUP BEST
  
group_best = str_split(tab_best$GROUP,pattern = "")[[1]]

load("../Robjects/MOD.BEST.ALLOTMENT.rdata")

plot(mod.best, hull = FALSE, ellipse = TRUE, main=c(paste("Axis1Axis2, PERMANOVA p-value = ",tab_best$betadiv_pvaleur,sep="")))

load("../Robjects/PHY.RAREFIED.BEST.ALLOTMENT.PLOT.rdata")


al1.best<-ggplot(phy.rarefied.best.plot, aes(x=Group, y=Richness, color=Group)) +
	  geom_boxplot() + geom_point(size = 2) + ggtitle(paste("Chao1, p-value = ",round(tab_best$richness_pvaleur,3),sep="")) + Theme_publi
	al2.best<-ggplot(phy.rarefied.best.plot, aes(x=Group, y=Evenness, color=Group)) +
	  geom_boxplot() + geom_point(size = 2) + ggtitle(paste("Shannon, p-value = ",round(tab_best$eveness_pvaleur,3),sep="")) + Theme_publi

cpubli_best <- cowplot::plot_grid(al1.best, al2.best, nrow = 1, ncol = 2, scale = .9, label_size = 28)
cpubli_best


```


# HOW TO FIND THE BEST DESIGN ?

Ranking score is a trustable metric to get homogenous group. The naïve way to find the best design would be to compute all possible combinations and retain only the minimal ranking score. 

Base on out first simulation, the computation time is exactly 60 minutes for 100 000 combinations.

We then simulate a few scenarios to estimate computation time for various design. 

```{r, echo=FALSE , warning=FALSE}

a <- comboGroupsCount(10, 2)
b <- comboGroupsCount(12, 2)
c <- comboGroupsCount(12, 3)
d <- comboGroupsCount(20, 2)
e <- comboGroupsCount(20, 4)
f <- comboGroupsCount(24, 2)
g <- comboGroupsCount(24, 3)
h <- comboGroupsCount(27, 3)
i <- comboGroupsCount(28,4)
j <- comboGroupsCount(30, 2)
k <- comboGroupsCount(30, 3)

sa <-  a*3600/100000
sb <-  b*3600/100000
sc <-  c*3600/100000
sd <-  d*3600/100000
se <-  e*3600/100000
sf <-  f*3600/100000
sg <-  g*3600/100000
sh <-  h*3600/100000
si <-  i*3600/100000
sj <-  j*3600/100000
sk <-  k*3600/100000

ta <- sa/3600/24/365.25
tb <- sb/3600/24/365.25
tc <- sc/3600/24/365.25
td <- sd/3600/24/365.25
te <- se/3600/24/365.25
tf <- sf/3600/24/365.25
tg <- sg/3600/24/365.25
th <- sh/3600/24/365.25
ti <- si/3600/24/365.25
tj <- sj/3600/24/365.25
tk <- sk/3600/24/365.25

df.time = data.frame(
  n_animals = c(10,12,12,20,20,24,24,27,28,30,30),
 n_groups = c(2,2,3,2,4,2,3,3,4,2,3),
n_possible_design  = c(a,b,c,d,e,f,g,h,i,j,k),
  CPU_time_sec = c(sa,sb,sc,sd,se,sf,sg,sh,si,sj,sk),
  CPU_time_year = c(ta,tb,tc,td,te,tf,tg,th,ti,tj,tk)
   )
colnames(df.time) = c("n animals","n groups","n possible designs","CPU time (sec)", "CPU time (year)")


df.time


```

There is no general formula for partitioning a vector into groups of equal size. 

The number of possible ways to partition a vector depends on the length of the vector and the size of the groups. 

As an illustration, The formula for calculating the number of ways to partition a vector into groups of equal size 2  is given by the following expression:

n! / (n / k)! * k^(n / k)

where n is the number of elements in the vector, k is the size of each group, and ! the factorial function.

Another way to approach this problem is to use generating functions, like the one propose by the R package RcppAlgos [ref]. 
 

For the experiment generated in the "constat", it is of course possible to generate all the combinations in 1 hours and get the best result. But for more complex design, it becomes quickly impossible to wait for the generation of all results and get an exact best match.


We then develop an algorithm based on HillDiv diversity measurements and anticlustering technics to obtain a optimal results in a very quick way.


# AN OPTIMAL DESIGN TO REDUCE EXPERIMENTAL BIASES

## ALGORITHM VALIDATION ON ALLOTMENT RESULTS

```{r, echo=FALSE, warning=FALSE}


load("../Robjects/RESULT.ALL.DIST.ANTICLUST.ALLOTMENT.400IT.rdata")

#objective : find the group into the whole possible design dataframe :
result.all.dist.at$GROUP2 = result.all.dist.at$GROUP
result.all.dist.at$GROUP2 = str_replace_all(string = result.all.dist.at$GROUP2,pattern = "2",replacement = "3")
result.all.dist.at$GROUP2 = str_replace_all(string = result.all.dist.at$GROUP2,pattern = "1",replacement = "2")
result.all.dist.at$GROUP2 = str_replace_all(string = result.all.dist.at$GROUP2,pattern = "3",replacement = "1")

result.all.dist.at$GROUP_FINAL = NA
for(i in 1:length(result.all.dist.at$COLUMN)){
  
  res = result.all.dist.at[i,]
  if( res$GROUP %in%  tab_stats_combinaison_ALL_WUNIFRAC$GROUP){
    result.all.dist.at$GROUP_FINAL[i] = res$GROUP
  }else{
    if( res$GROUP2 %in% tab_stats_combinaison_ALL_WUNIFRAC$GROUP){
    result.all.dist.at$GROUP_FINAL[i] = res$GROUP2
  }
    
  }
}

#minimum ranking score = best design
result.best = result.all.dist.at[ result.all.dist.at$METRIC_PVALUE == min(result.all.dist.at$METRIC_PVALUE),]

tab_stats_result_allotement= tab_stats_combinaison_ALL_WUNIFRAC[tab_stats_combinaison_ALL_WUNIFRAC$GROUP %in% result.all.dist.at$GROUP_FINAL,]

#tab_stats_result_allotement= tab_stats_combinaison_ALL[tab_stats_combinaison_ALL$GROUP %in% result.best$GROUP_FINAL,]
tab_stats_result_allotement=tab_stats_result_allotement[tab_stats_result_allotement$METRIC_PVALUE == min(tab_stats_result_allotement$METRIC_PVALUE),]


#generate same graphic from previously but identifying suggested design 
compare_mean <- tab_stats_combinaison_sorted %>%
        dplyr::group_by(group) %>%
        dplyr::summarise(Mean = mean(METRIC_PVALUE))

compare_mean <- compare_mean %>% 
   dplyr::add_row(group = "SUGGESTED DESIGN", Mean=tab_stats_result_allotement$METRIC_PVALUE)



output_histo_allotement_opti <- ggplot(tab_stats_combinaison_sorted, aes(x = METRIC_PVALUE, fill = group)) +
        geom_histogram(binwidth = 20, alpha = .5, position = "identity")  + scale_fill_manual(values=c("red", "green", "orange")) +
        geom_vline(data = compare_mean[4,], aes(xintercept = Mean,color="SUGGESTED DESIGN"),linetype = "dashed", size = 1) + labs(x = "RANKING SCORE") + Theme_publi 

output_histo_allotement_opti

###GROUP OPTI BEST
tab_best = tab_stats_result_allotement

load("../Robjects/MOD.ALLOTMENT.rdata")

plot(mod.allotment, hull = FALSE, ellipse = TRUE, main=c(paste("Axis1Axis2, PERMANOVA p-value = ",tab_best$betadiv_pvaleur,sep="")))

load("../Robjects/PHY.RAREFIED.OPTI.BEST.ALLOTMENT.PLOT.rdata")

#alpha div
al1<-ggplot(phy.sam.data, aes(x=Group, y=Richness, color=Group)) +
	  geom_boxplot() + geom_point(size = 2) + ggtitle(paste("Chao1, p-value = ",round(tab_best$richness_pvaleur,3),sep=""))+ Theme_publi #+  theme( panel.background = element_rect(fill = "white", colour = "black"), panel.grid.major=element_line(colour = "grey"))
	al2<-ggplot(phy.sam.data, aes(x=Group, y=Evenness, color=Group)) +
	  geom_boxplot() + geom_point(size = 2) + ggtitle(paste("Shannon, p-value = ",round(tab_best$eveness_pvaleur,3),sep="")) + Theme_publi#+  theme( panel.background = element_rect(fill = "white", colour = "black"), panel.grid.major=element_line(colour = "grey"))
	cpubli.allotment <- cowplot::plot_grid(al1, al2,nrow = 1, ncol = 2, scale = .9, label_size = 28)
cpubli.allotment


```



## ALGORITHM VALIDATION ON THIRD PARRY DATA (Neuman et al.)

###  HISTOGRAM - 90000 RANDOMLY SIMULATING 90000 DESIGNS

```{r, echo=FALSE,warning=FALSE}
#load phyloseq object from SRA experiement already rarefied at 32629
load("../Robjects/OUTPUT_PHYLOSEQ_RAREFIED_R_OBJECT_SRA.rdata")

#load raw phyloseq object
load("../Robjects/OUTPUT_PHYLOSEQ_R_OBJECT_SRA.rdata")

#load 90000 random sampling designs - tab_stats_combinaison_SRA_ALL
load("../Robjects/OUTPUT_TAB_STATS_COMBINATION_SRA_ALL_90000_WUNIFRAC_PUBLICATION_V1.rdata",verbose=TRUE)

#estimate alpha diversiyt metrics
rich = estimate_richness(phy.rarefied, measures = c("Chao1","InvSimpson"))

Richness= rich[,colnames(rich) == "Chao1"]
Evenness =  rich[,colnames(rich) == "InvSimpson"]
meta <- cbind(normalize_custom(Richness), normalize_custom(Evenness))
rownames(meta) = rownames(rich)
phy.rarefied@sam_data$Richness = Richness

phy.rarefied@sam_data$Evenness = Evenness

phy.sra = phy.rarefied


load("../Robjects/RESULT.ALL.DIST.ANTICLUST.SRA.400IT.rdata")


result.sra= tab_stats_combinaison_SRA_ALL[length(tab_stats_combinaison_SRA_ALL$COLUMN),]

tab_stats_combinaison_SRA_sorted =tab_stats_combinaison_SRA_ALL[ order(tab_stats_combinaison_SRA_ALL$METRIC_PVALUE),]

quantile_metric= quantile(tab_stats_combinaison_SRA_sorted$METRIC_PVALUE, probs = c(0.05,0.95))
tab_stats_combinaison_SRA_sorted$group = "SUB-OPTIMAL DESIGN"
tab_stats_combinaison_SRA_sorted$group[tab_stats_combinaison_SRA_sorted$METRIC_PVALUE <= as.numeric(quantile_metric[1]) ] = "OPTIMAL DESIGN"
tab_stats_combinaison_SRA_sorted$group[tab_stats_combinaison_SRA_sorted$METRIC_PVALUE >= as.numeric(quantile_metric[2]) ] = "HIGH-RISK DESIGN"

result.best.sra =  result.all.dist.sra[ result.all.dist.sra$METRIC_PVALUE == min(result.all.dist.sra$METRIC_PVALUE),]
 load("../Robjects/OUTPUT_TAB_STATS_COMBINATION_SRA_CUSTOM_PUBLICATION_V1.rdata") 
  
tab_stats_combinaison_SRA_sorted =tab_stats_combinaison_SRA_CUSTOM[ order(tab_stats_combinaison_SRA_CUSTOM$METRIC_PVALUE),]

quantile_metric= quantile(tab_stats_combinaison_SRA_sorted$METRIC_PVALUE, probs = c(0.05,0.95))
tab_stats_combinaison_SRA_sorted$group = "SUB-OPTIMAL DESIGN"
tab_stats_combinaison_SRA_sorted$group[tab_stats_combinaison_SRA_sorted$METRIC_PVALUE <= as.numeric(quantile_metric[1]) ] = "OPTIMAL DESIGN"
tab_stats_combinaison_SRA_sorted$group[tab_stats_combinaison_SRA_sorted$METRIC_PVALUE >= as.numeric(quantile_metric[2]) ] = "HIGH-RISK DESIGN"

tab_stats_result_sra= tab_stats_combinaison_SRA_sorted[tab_stats_combinaison_SRA_sorted$GROUP %in% result.best.sra$GROUP,]
tab_stats_study_sra= tab_stats_combinaison_SRA_sorted[tab_stats_combinaison_SRA_sorted$GROUP %in% result.sra$GROUP,]

#generate figure - including SRA study design and suggested design by Allomento

compare_mean <- tab_stats_combinaison_SRA_sorted %>%
        dplyr::group_by(group) %>%
        dplyr::summarise(Mean = mean(METRIC_PVALUE))


compare_mean <- compare_mean %>% 
   add_row(group = "SUGGESTED DESIGN", Mean=tab_stats_result_sra$METRIC_PVALUE)
compare_mean <- compare_mean %>% 
   add_row(group = "STUDY DESIGN", Mean=tab_stats_study_sra$METRIC_PVALUE)

output_histo_sra <- ggplot(tab_stats_combinaison_SRA_sorted, aes(x = METRIC_PVALUE, fill = group)) +
        geom_histogram(binwidth = 20, alpha = .5, position = "identity")  + scale_fill_manual(values=c("red", "green", "orange")) +
        geom_vline(data = compare_mean[4,], aes(xintercept = Mean,color="SUGGESTED DESIGN"),linetype = "dashed", size = 1) +
        geom_vline(data = compare_mean[5,], aes(xintercept = Mean,color="STUDY DESIGN"),linetype = "dashed", size = 1) + labs(x = "RANKING SCORE")  + Theme_publi 
output_histo_sra 
```

### ALPHA AND BETA DIVERSITIES VALIDATION ON BACT-TO-BATCH SUGGESTED DESIGN

```{r allotmento design, echo=FALSE, warning=FALSE}

#Visualize SRA design and study design
  phy.asv = data.frame(phy.rarefied@otu_table)
  phy.tree = phy.rarefied@phy_tree
  unifracs <- GUniFrac(t(phy.asv), phy.tree, alpha=c(0, 0.5, 1))$unifracs
  Dist_BASE_M <- unifracs[, , "d_1"]
  
#suggested design  
group_best = str_split(result.best.sra$GROUP,pattern = "")[[1]]
phy.rarefied.sra.opti.best=phy.rarefied
phy.rarefied.sra.opti.best@sam_data$BLOCK  = group_best


#betadiv
groups <- factor(phy.rarefied.sra.opti.best@sam_data$BLOCK, labels = c("TPC","PBS","Tuftsin"))
mod.sra.opti.best <- betadisper(as.dist(Dist_BASE_M), groups,)
plot(mod.sra.opti.best, hull = FALSE, ellipse = TRUE, main=c(paste("Axis1Axis2, PERMANOVA p-value = ",result.best.sra$betadiv_pvaleur,sep="")))

phy.rarefied.sra.opti.best@sam_data$Group[ phy.rarefied.sra.opti.best@sam_data$BLOCK == 1] = "TPC"
phy.rarefied.sra.opti.best@sam_data$Group[ phy.rarefied.sra.opti.best@sam_data$BLOCK == 2] = "PBS"
phy.rarefied.sra.opti.best@sam_data$Group[ phy.rarefied.sra.opti.best@sam_data$BLOCK == 3] = "Tuftsin"

#alpha div
al1<-ggplot(phy.rarefied.sra.opti.best@sam_data[!is.na(phy.rarefied.sra.opti.best@sam_data$BLOCK)], aes(x=Group, y=Richness, color=Group)) +
	  geom_boxplot() + geom_point(size = 2) + ggtitle(paste("Chao1, p-value = ",round(result.best.sra$richness_pvaleur,3),sep="")) + Theme_publi 
	al2<-ggplot(phy.rarefied.sra.opti.best@sam_data[!is.na(phy.rarefied.sra.opti.best@sam_data$BLOCK)], aes(x=Group, y=Evenness, color=Group)) +
	  geom_boxplot() + geom_point(size = 2) + ggtitle(paste("Shannon, p-value = ",result.best.sra$eveness_pvaleur,sep="")) + Theme_publi
	cpubli <- cowplot::plot_grid(al1, al2,nrow = 1, ncol = 2, scale = .9, label_size = 28)
cpubli

```

### ALPHA AND BETA DIVERSITIES VALIDATON ON STUDY DESIGN

```{r study design, echo=FALSE, warning=FALSE}

#sra study design  
group_study = str_split(tab_stats_study_sra$GROUP,pattern = "")[[1]]
phy.rarefied.sra.opti.study=phy.rarefied
phy.rarefied.sra.opti.study@sam_data$BLOCK  = group_study


#betadiv
groups <- factor(phy.rarefied.sra.opti.study@sam_data$BLOCK, labels = c("TPC","PBS","Tuftsin"))
mod <- betadisper(as.dist(Dist_BASE_M), groups,)
plot(mod, hull = FALSE, ellipse = TRUE, main=c(paste("Axis1Axis2, PERMANOVA p-value = ",tab_stats_study_sra$betadiv_pvaleur,sep="")))

phy.rarefied.sra.opti.study@sam_data$Group[ phy.rarefied.sra.opti.study@sam_data$BLOCK == 1] = "TPC"
phy.rarefied.sra.opti.study@sam_data$Group[ phy.rarefied.sra.opti.study@sam_data$BLOCK == 2] = "PBS"
phy.rarefied.sra.opti.study@sam_data$Group[ phy.rarefied.sra.opti.study@sam_data$BLOCK == 3] = "Tuftsin"

#alpha div
al1<-ggplot(phy.rarefied.sra.opti.study@sam_data[!is.na(phy.rarefied.sra.opti.study@sam_data$BLOCK)], aes(x=Group, y=Richness, color=Group)) +
	  geom_boxplot() + geom_point(size = 2) + ggtitle(paste("Chao1, p-value = ",round(tab_stats_study_sra$richness_pvaleur,3),sep="")) + Theme_publi 
	al2<-ggplot(phy.rarefied.sra.opti.study@sam_data[!is.na(phy.rarefied.sra.opti.study@sam_data$BLOCK)], aes(x=Group, y=Evenness, color=Group)) +
	  geom_boxplot() + geom_point(size = 2) + ggtitle(paste("Shannon, p-value = ",round(tab_stats_study_sra$eveness_pvaleur,3),sep="")) +  Theme_publi
	c <- cowplot::plot_grid(al1, al2,nrow = 1, ncol = 2, scale = .9, label_size = 14)
c


```

